---
title: "Protected Species Interaction Modeling"
format: html
editor: visual
---

```{r}
# Packages used
library(tidyverse)
library(modelr)
library(ggplot2)
library(janitor)
library(lubridate)
library(tidybayes)
library(rstanarm)
library(bayesrules)
```

# 1. Introduction

This analysis answers:

> **How does hook/set effort and species type correlate with the likelihood and intensity of protected species interactions?**

Outcomes:

-   Probability of at least one protected species interaction (`protected_interaction`, 0/1)

-   Expected number of protected species interactions (`protected_count`, integer)

Data source: MRIP catch CSV provided by project partners.

```{r}
marine_catch <- read_csv("ps_2024_csv/catch_20241.csv") 

marine_catch
```

```{r}
marine_catch <- marine_catch %>%
janitor::clean_names()
```

```{r}
select(marine_catch, common, tot_cat, wgt_ab1, tot_len, landing, mode_fx, area_x, wave, year, st) %>%
slice_head(n=10)
```

```{r}
protected_species <- c(
"LOGGERHEAD", "GREEN TURTLE", "LEATHERBACK", "HAWKSBILL", # example turtle names
"DOLPHIN", "PORPOISE", # cetaceans (if recorded)
"SAND TIGER", "SAND TIGER SHARK", "BONNETHEAD", "DUSKY SHARK"
)
```

```{r}
marine_catch <- marine_catch %>%
mutate(
common_up = toupper(if_else(is.na(common), "", common))
)
```

```{r}
marine_catch <- marine_catch %>%
mutate(
protected_interaction = if_else(common_up %in% protected_species & !is.na(tot_cat) & tot_cat > 0, 1L, 0L),
protected_count = if_else(common_up %in% protected_species, if_else(is.na(tot_cat), 0L, as.integer(tot_cat)), 0L)
)
```

```{r}
marine_catch <- marine_catch %>%
mutate(
species_group = case_when(
str_detect(common_up, "TURTLE|SEA TURTLE|LOGGERHEAD|GREEN|LEATHERBACK|HAWKSBILL") ~ "turtle",
str_detect(common_up, "SHARK|TIGER|BONNETHEAD|DUSKY|SAND") ~ "shark",
str_detect(common_up, "DOLPHIN|PORPOISE") ~ "cetacean",
common_up == "" ~ "unknown",
TRUE ~ "other_fish"
),
wave = as.factor(wave),
area_x = as.factor(area_x),
mode_fx = as.factor(mode_fx),
year = as.integer(year)
)
```

```{r}
marine_catch %>% select(common, common_up, species_group, protected_interaction, protected_count, tot_cat, wave, area_x, mode_fx) %>% slice_head(n=12)
```

```{r}
#Count total protected interactions (weighted & unweighted)
summary_df <- marine_catch %>%
summarise(
total_records = n(),
total_protected_records = sum(protected_interaction, na.rm = TRUE),
total_protected_individuals = sum(protected_count, na.rm = TRUE)
)

summary_df
```

```{r}
#Protected interactions by species group
marine_catch %>%
group_by(species_group) %>%
summarise(
records = n(),
protected_records = sum(protected_interaction, na.rm = TRUE),
protected_individuals = sum(protected_count, na.rm = TRUE)
) %>%
arrange(desc(protected_individuals)) %>%
ggplot(aes(x = fct_reorder(species_group, protected_individuals), y = protected_individuals), color = protected_individuals) +
geom_col() +
coord_flip() +
labs(title = "Protected individuals by species group",
x = "", y = "Protected individuals (count)")
```

```{r}
marine_catch %>%
group_by(wave) %>%
summarise(protected_records = sum(protected_interaction, na.rm = TRUE)) %>%
ggplot(aes(x = wave, y = protected_records)) +
geom_col() +
labs(title = "Protected interactions by sampling wave (season)", x = "Wave", y = "Protected interactions")
```

```{r}
marine_catch %>%
group_by(mode_fx, area_x) %>%
summarise(protected = sum(protected_interaction, na.rm = TRUE)) %>%
ggplot(aes(x = mode_fx, y = protected, fill = area_x)) +
geom_col(position = "dodge") +
labs(title = "Protected interactions by fishing mode and area", x = "Mode", y = "Protected interactions")
```

```{r}
marine_catch %>%
summarise(mean_tot_cat = mean(tot_cat, na.rm = TRUE),
median_tot_cat = median(tot_cat, na.rm = TRUE),
pct_with_catch = mean(tot_cat > 0, na.rm = TRUE))
```

```{r}
use_rstanarm <- requireNamespace("rstanarm", quietly = TRUE)
use_brms <- requireNamespace("brms", quietly = TRUE)
use_arm <- requireNamespace("arm", quietly = TRUE)

message("Packages available: rstanarm=", use_rstanarm, " brms=", use_brms, " arm=", use_arm)
```

```{r}
form_logit <- as.formula("protected_interaction ~ mode_fx + area_x + wave + species_group + year")
form_count <- as.formula("protected_count ~ mode_fx + area_x + wave + species_group + year")
```

```{r}
# Collapse to trip-level data
trip_level <- marine_catch %>%
  group_by(id_code) %>%
  summarise(
    protected_hit = as.integer(any(protected_interaction == 1)),
    total_catch = sum(tot_cat, na.rm = TRUE),
    mode_fx = first(mode_fx),
    area_x = first(area_x),
    wave = first(wave),
    year = first(year)
  )

# Logistic regression (Bernoulli likelihood)
fit_logit <- stan_glm(
  protected_hit ~ total_catch + mode_fx + area_x,
  data = trip_level,
  family = binomial(link = "logit"),
  prior = normal(0, 1),
  prior_intercept = normal(0, 2),
  seed = 123
)

fit_logit
```

```{r}
logit_draws <- tidy_draws(fit_logit)

logit_draws %>%
  median_qi()
```

```{r}
exp(posterior_interval(fit_logit, prob = 0.80))
```

```{r}
post_draws <- posterior_predict(fit_logit, newdata = trip_level, draws = 300)

```

```{r}


trip_level <- trip_level %>% mutate(.row = row_number())


post_long <- as.data.frame(t(post_draws)) %>%  # transpose so rows = rows of trip_level
  mutate(.row = row_number()) %>%
  pivot_longer(cols = - .row, names_to = "draw", values_to = "pred") %>%
  mutate(draw = as.integer(draw))

post_long <- post_long %>%
  left_join(trip_level, by = ".row")

posterior_summary <- post_long %>%
  group_by(total_catch) %>%
  summarise(
    mean_prob = mean(pred),
    lower = quantile(pred, 0.025),
    upper = quantile(pred, 0.975),
    .groups = "drop"
  )

ggplot(posterior_summary, aes(x = total_catch, y = mean_prob)) +
  geom_line(color = "steelblue") +
  labs(
    x = "Total Catch",
    y = "Posterior Probability of Protected Species Interaction",
    title = "Predicted Probability with 95% Credible Interval"
  )


  
```

```{r}
ggplot(posterior_summary, aes(x = total_catch, y = mean_prob)) +
  geom_line(color = "steelblue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(
    x = "Total Catch",
    y = "Posterior Probability of Protected Species Interaction",
    title = "Predicted Probability with 95% Credible Interval"
  )
```

```{r}

```
